---
output: github_document
always_allow_html: yes
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE, cache = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  cache = FALSE,
  asciicast_at = "end",
  asciicast_theme = if (Sys.getenv("IN_PKGDOWN") == "true") "pkgdown" else "readme"
)
asciicast::init_knitr_engine(
  startup = quote({
    library(cli)
    library(vlightr)
    options(cli.num_colors = cli::truecolor)
    set.seed(1)
  }),
  echo = TRUE,
  echo_input = FALSE
)
```

# vlightr

<!-- badges: start -->

<!-- badges: end -->

vlightr is a package for conditionally formatting vectors. You can `highlight()`
a vector to style, colour, or otherwise re-format it's elements when it is 
printed (via `print()`) or formatted (via `format()`).

## Installation

You can install the development version of vlightr from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("EthanSansom/vlightr")
```

## Features

```{r load-pkg}
library(vlightr)
```

#### Conditionally format vector elements

```{asciicast red-nas1}
x <- c(1L, 0L, NA, 1L, 0L)
print(x)
```

```{asciicast red-nas2}
x_hl <- vlightr::highlight(x, is.na, vlightr::colour("red"))
print(x_hl)
```

#### Manipulate highlighted data

Highlighted vectors play well with other highlighted vectors. The shorthand 
`hl()` is useful for quick highlighting.

```{asciicast operation}
# Arithmetic
x_hl + vlightr::hl(10L)
```

```{asciicast coerce}
# Coercion
c(x_hl, vlightr::hl(c(-1.5, NA)))
```

#### Apply multiple conditional formats

```{asciicast multiple-formats}
dummies <- vlightr::highlight(
  x = x,
  conditions = list(
    is.na, 
    ~ .x == 1, 
    ~ .x == 0
  ),
  formatters = list(
    vlightr::colour("red"), 
    ~ paste(.x, "[Yes]"),
    ~ paste(.x, "[No]")
  )
)
dummies
```

```{asciicast unhighlight}
# The underlying data is unchanged
vlightr::un_highlight(dummies)
```

#### `dplyr::case_when()` style syntax 

```{asciicast case-syntax}
bad_words <- c("darn", "gosh")
affirmations <- c("good job", "way-to-go")
message <- c("hey", "good job", "but", "darn", "please", "don't", "say", "gosh")

vlightr::highlight_case(
  message,
  .x %in% bad_words ~ \(x) strrep("X", nchar(x)),
  .x %in% affirmations ~ \(x) toupper(x),
  TRUE ~ cli::style_italic
)
```

#### Highlight vectors from other packages

```{asciicast generic}
library(lubridate, warn.conflicts = FALSE)

today <- ymd("2020-01-01")
meeting_times <- interval(
  today + hours(c(9, 11, 16)), 
  today + hours(c(10, 13, 17))
)
lunch_break <- interval(today + hours(12), today + hours(13))
is_during_lunch <- function(x) int_overlaps(x, lunch_break)

vlightr::highlight(meeting_times, is_during_lunch, cli::col_magenta)
```
