---
output: github_document
always_allow_html: yes
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE, cache = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  cache = TRUE,
  asciicast_at = "end",
  asciicast_theme = if (Sys.getenv("IN_PKGDOWN") == "true") "pkgdown" else "readme"
)
asciicast::init_knitr_engine(
  startup = quote({
    library(cli)
    library(vlightr)
    options(cli.num_colors = cli::truecolor)
    set.seed(1)
  }),
  echo = TRUE,
  echo_input = FALSE
)
```

# vlightr

<!-- badges: start -->

<!-- badges: end -->

vlightr is a package for

## Installation

You can install the development version of vlightr from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("EthanSansom/vlightr")
```

## Features

```{r}
library(vlightr)
```

```{asciicast setup, show = FALSE}
library(vlightr)
```

Conditionally format vector elements.

```{asciicast red-nas}
x <- c(1L, 0L, NA, 1L, 0L)
x_hl <- vlightr::highlight(x, is.na, vlightr::colour("red"))
x_hl
```

Highlighted vectors play well with other highlighted vectors of the same type.
The shorthand `hl()` is useful for quick highlighting.

```{asciicast operation}
x_hl + vlightr::hl(10L)
```

Coercion works as well.

```{asciicast coerce}
c(x_hl, vlightr::hl(c(-1.5, NA)))
```

Apply multiple conditional formats.

```{asciicast multiple-formats}
vlightr::highlight(
  x = x,
  conditions = list(
    is.na, 
    ~ .x == 1, 
    ~ .x == 0
  ),
  formatters = list(
    vlightr::colour("red"), 
    ~ paste(.x, "[Yes]"),
    ~ paste(.x, "[No]")
  )
)
```

Use `dplyr::case_when()` style syntax for multiple conditional formats.

```{asciicast case-syntax}
vlightr::highlight_case(
  x = c(10, NA, -11, , 2, -1, 8),
  is.na(.x) ~ \(x) cli::col_green("Don't Worry"),
  .x < 0 ~ vlightr::bg("bright_blue")
)
```

You'll still need to render `README.Rmd` regularly, to keep `README.md` up-to-date. `devtools::build_readme()` is handy for this. Hey!
