---
title: vlightr
output:
  github_document:
always_allow_html: yes
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE, cache = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  cache = FALSE,
  asciicast_at = "end",
  # This tells `asciicast` to generate HTML for the website, SVG for GitHub
  asciicast_theme = if (Sys.getenv("IN_PKGDOWN") == "true") "pkgdown" else "readme"
)
asciicast::init_knitr_engine(
  startup = quote({
    library(cli)
    library(vlightr)
    library(dplyr, warn.conflicts = FALSE)
    library(ivs)
    options(
      vlightr.default_test = vlightr::false,
      # This shows up better on github light and dark mode
      vlightr.default_formatter = vlightr::color("goldenrod"),
      cli.num_colors = cli::truecolor
    )
    set.seed(1)
  }),
  echo = TRUE,
  echo_input = FALSE
)
```

<!-- badges: start -->

<!-- badges: end -->

{vlightr} (read, "vector-highlighter") makes identifying key elements of vectors easier. It implements a `<vlightr_highlight>` vector superclass which enhances the `format()` and `print()` method of generic vectors, allowing vector elements to be conditionally formatted.

## Installation

⚠️ This package is still **under construction**. ⚠️

You can install the development version of vlightr from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("EthanSansom/vlightr")
```

## Features

```{r load-pkg}
library(vlightr)
library(dplyr, warn.conflicts = FALSE)
```

Want to identify an element of a vector? Highlight it with `highlight()`.

```{asciicast max-values}
x <- c(1, 8, 12, 4, 2)
# maximum_hl <- vlightr::highlight(x, .t = ~ .x == max(.x), .f = ~ vlightr::bg("br_yellow"))
maximum_hl <- vlightr::highlight(x, .t = ~ .x == max(.x))
print(maximum_hl)
```

Highlighted elements change as the highlighted vector changes, so you won’t lose them.

```{asciicast sort-demo}
# `hl()` is shorthand for `highlight()`
sort(maximum_hl + vlightr::hl(10))
```

Highlighted vectors can be used as tibble::tibble() columns too.

```{asciicast tibble}
iris |>
  as_tibble() |>
  mutate(
    species = vlightr::highlight_mult(
      Species,
      "setosa" ~ vlightr::color("purple"),
      "versicolor" ~ vlightr::color("violet"),
      "virginica" ~ vlightr::color("pink")
    )
  ) |>
  group_by(species) |>
  summarize(
    avg_petal_length = vlightr::highlight(mean(Petal.Length), ~ .x == max(.x)),
    avg_sepal_width = vlightr::highlight(mean(Sepal.Width), ~ .x == max(.x))
  ) |>
  ungroup()
```

Are you (or your boss) having a hard time finding that row you're looking for? Use `templight()` to temporarily highlight a vector by index instead of using a test function.

```{asciicast templight}
mtcars |>
	as_tibble(rownames = "make") |>
	mutate(across(everything(), ~ vlightr::templight(.x, make == "Datsun 710"))) |>
	select(make, mpg, disp, vs) |>
  head(5)
```

You can apply multiple conditional formats to a vector using `highlight_mult()`. The left-hand-side is you a test function or a literal value that you want to match and the right-hand-side is your formatter function.

```{asciicast highlight-mult}
indicator <- vlightr::highlight_mult(
	c(1, 0, 1, 0, 0, NA, 5),
	is.na ~ vlightr::color("red"),
	0 ~ vlightr::label("No"),
	1 ~ vlightr::label("Yes"),
	!(.x %in% c(NA, 0, 1)) ~ paste(.x, "[?]")
)
print(indicator)
```

Simplify the code above using `highligh_case()`, which provides a `dplyr::case_when()` style interface and conditionally formats elements using at most one formatter.

```{asciicast highlight-case}
indicator <- vlightr::highlight_case(
	c(1, 0, 1, 0, 0, NA, 5),
	is.na ~ vlightr::color("red"),
	0 ~ vlightr::label("No"),
	1 ~ vlightr::label("Yes"),
	vlightr::true ~ paste(.x, "[?]") # Use `true()` to provide a default formatter
)
print(indicator)
```

If you want to re-use a highlight, make a `highlighter()` and friends.

```{asciicast highlighter}
indicator_highlighter <- vlightr::highlighter_case(
	is.na ~ vlightr::color("red"),
	0 ~ vlightr::label("No"),
	1 ~ vlightr::label("Yes"),
	vlightr::true ~ paste(.x, "[?]")
)
indicator_highlighter(c(0, 1, NA, -9))
```

## Inspiration

The development of {vlightr} relied heavily on the following pacakges:

-   [purrr](https://purrr.tidyverse.org/), which inspired the user interface of `highlight()`
-   [dplyr](https://dplyr.tidyverse.org/), whose `case_when()` inspired `highlight_case()`
-   [magrittr](https://magrittr.tidyverse.org/), whose `%>%` made the highlight pipe `%hl>%` possible
-   [vctrs](https://vctrs.r-lib.org/index.html), for making S3 vectors easy to work with
-   [usethis](https://usethis.r-lib.org/) and [pkgdown](https://pkgdown.r-lib.org/), which make every step of R-package and R-package-website development easier

This package is also heavily inspired by the [ivs](https://github.com/DavisVaughan/ivs) package, which (also powered by [vctrs](https://vctrs.r-lib.org/index.html)) implements generic right-open intervals defined by a pair of parallel start and end vectors.

As a testament to the genericity of the `<ivs_iv>`, here is an ill-advised but perfectly legal interval vector.

```{asciicast ivs-inspo1}
library(ivs)

# Prepare highlighted numeric start and end vectors
starts <- vlightr::highlight(-3:2, ~ .x %% 2 == 0, ~ label("Even"))
ends <- vlightr::highlight(c(-2, -1, 2, 5, 7, 8), ~ .x > 0, ~ paste0("+", .x))

# Make an iv() with highlighted `starts` and `ends`
iv(starts, ends)
```

The interval vector can be manipulated as you'd expect.

```{asciicast ivs-inspo2}
iv_groups(iv(starts, ends))
```

And the interval vector can itself be highlighted.

```{asciicast ivs-inspo3}
vlightr::highlight(
  iv(starts, ends), 
  ~ (iv_end(.x) - iv_start(.x)) > hl(1),
  vlightr::color("goldenrod")
)
```
