---
title: vlightr
output:
  github_document:
always_allow_html: yes
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE, cache = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  cache = FALSE,
  asciicast_at = "end",
  # This tells `asciicast` to generate HTML for the website, SVG for GitHub
  asciicast_theme = if (Sys.getenv("IN_PKGDOWN") == "true") "pkgdown" else "readme"
)
asciicast::init_knitr_engine(
  startup = quote({
    library(cli)
    library(vlightr)
    options(cli.num_colors = cli::truecolor)
    set.seed(1)
  }),
  echo = TRUE,
  echo_input = FALSE
)
```

<!-- badges: start -->

<!-- badges: end -->

vlightr is a package for conditionally formatting vectors. You can `highlight()` a vector to style, colour, or otherwise re-format it's elements when the vector is printed or formatted. Highlights are:

-   **Persistent**, meaning highlighted vectors can be manipulated (with minimal legwork) while maintaining their custom formatting.
-   **Generic**, meaning that many vectors you'll encounter in the wild and all base atomic vectors (`"logical"`, `"integer"`, `"numeric"`, `"complex"`, `"character"` and `"raw"`) can be highlighted.

## Installation

⚠️ This package is still **under construction**. ⚠️

You can install the development version of vlightr from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("EthanSansom/vlightr")
```

## Features

```{r load-pkg}
library(vlightr)
library(lubridate, warn.conflicts = FALSE)
```

Quickly customize how a vector is printed.

```{asciicast june1}
dates <- lubridate::ymd("20240619") + lubridate::weeks(-3:2)

format_roygbiv1 <- function(x) {
  # vroygbi <- c("violet", "red", "orange", "br_yellow", "green", "blue", "purple")
  vroygbi <- c("violet", "red", "orange", "gold", "green", "blue", "purple")
  crayons <- lapply(vroygbi, vlightr::color)
  letters <- strsplit(as.character(x), "")[[1]]
  letters |>
    mapply(seq_along(letters), FUN = \(x, i) crayons[[i %% 7 + 1]](x)) |>
    paste(collapse = "")
}
format_roygbiv <- function(x) sapply(x, format_roygbiv1)

dates <- vlightr::highlight(dates, ~ lubridate::month(.x) == 6, format_roygbiv)
print(dates)
```

And how a column is displayed.

```{asciicast june2}
tibble::tibble(
  month = lubridate::month(vlightr::ul(dates), label = TRUE),
  date = dates
)
```

### Highlight, Un-Highlight, Re-Highlight

Highlighted vectors can't be implicitly coerced or converted to another vector type. To use a function which expects the highlight's underlying type (ex. a date) first `un_highlight()` (AKA `ul()`) to expose the highlighted data and then `re_highlight()` (AKA `rl()`) to re-apply the conditional formatting.

```{asciicast rehighlight1}
dates |>
  vlightr::ul() |>
  lubridate::rollback() |>
  vlightr::rl(dates)
```

The "highlight-pipe" `%hl>%` wraps the magrittr `%>%` to do this automatically.

```{asciicast rehighlight2}
dates %hl>% lubridate::rollback()
```

### Manipulate Highlighted Data

Highlighted vectors are happy to be cast or coerced to other compatible highlights. The shorthand `hl()` is useful for quickly highlighting a vector for this purpose.

```{asciicast operation}
# Arithmetic
dates + vlightr::hl(17)
```

```{asciicast assign}
# Assignment
dates[[1]] <- vlightr::hl(NA)
dates
```

```{asciicast coerce}
# Coercion
c(dates, vlightr::hl(lubridate::ymd_hm("2020-06-01 12:00")))
```

### Multiple Conditional Formats

Multiple conditions and formatters can be specified by supplying lists
of functions or purrr-style lambdas. Elements of `x` for which `conditions[[i]](x)` 
is `TRUE` are formatted using the formatter function `formatters[[i]]`.

```{asciicast multiple-formats}
dummies <- vlightr::highlight(
  x = c(1L, 0L, NA, 1L, 0L),
  conditions = list(
    is.na, 
    ~ .x == 1, 
    ~ .x == 0
  ),
  formatters = list(
    vlightr::colour("red"), 
    ~ paste(.x, "[Yes]"),
    ~ paste(.x, "[No]")
  )
)
dummies
```

`highlight_case()` provides a `dplyr::case_when` inspired syntax for defining
multiple conditional formats.

```{asciicast case-syntax}
bad_words <- c("darn", "gosh")
affirmations <- c("good job", "way-to-go")
message <- c("hey", "good job", "but", "darn", "please", "don't", "say", "gosh")

vlightr::highlight_case(
  message,
  .x %in% bad_words ~ strrep("X", nchar(.x)),
  .x %in% affirmations ~ toupper(.x),
  TRUE ~ cli::style_italic
)
```

### Define Highlighter Functions

Re-use the formatting of any highlighted vector by converting in into a highlighter.

```{asciicast highlighter}
dummy_highlighter <- vlightr::as_highlighter(dummies)
dummy_highlighter(c(0, 1, NA))
```

### Highlight Arbitrary Vectors

Highlights are generic, meaning that S3 and S4 vector classes from other packages are highlight-able.

```{asciicast generic}
library(lubridate, warn.conflicts = FALSE)

today <- ymd("2020-01-01")
meeting_times <- interval(
  today + hours(c(9, 11, 16)), 
  today + hours(c(10, 13, 17))
)
lunch_break <- interval(today + hours(12), today + hours(13))
is_during_lunch <- function(x) int_overlaps(x, lunch_break)

vlightr::highlight(meeting_times, is_during_lunch, cli::col_magenta)
```

## Inspiration

This package is heavily inspired by the [ivs](https://github.com/DavisVaughan/ivs) package, which implements generic right-open intervals defined by a pair of parallel start and end vectors.

As a testament to the genericity of the `ivs::iv`, here is an ill-advised but perfectly legal interval vector.

```{r}
library(ivs)

starts <- highlight(-3:2, ~ .x %% 2 == 0, ~ paste(.x, "[Even]"))
ends <- highlight(c(-2, -1, 2, 5, 7, 8), ~ .x > 0, ~ paste0("+", .x))

# A totally legitimate `iv`
ivs::iv(starts, ends)
```

```{r}
# We can even manipulate it
ivs::iv_groups(ivs::iv(starts, ends))
```

```{r}
# Or highlight it...
highlight(
  ivs::iv(starts, ends), 
  ~ (ivs::iv_end(.x) - ivs::iv_start(.x)) > hl(1),
  ~ paste("{", .x, "}")
)
```
