---
title: "Highlighting and dplyr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{highlighting-and-dplyr}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

This vignette describes some functions which are designed for working
with highlighted vectors as columns in a `data.frame` or `tibble` and
provides some examples of manipulating highlights using `dplyr`
functions.

```{r load}
library(vlightr)
library(dplyr, warn.conflicts = FALSE)
```

```{r constants, include = FALSE}
n_droids <- sum(starwars$species == "Droid", na.rm = TRUE)
n_rows <- nrow(starwars)
```

Before we begin, let's define a new formatter function `emph()` which adds a
prefix `left` and a suffix `right` to the text `x` and colors the text 
background the color `col`. If you are reading this vignette in R (say within 
the RStudio Help pane) then you likely won't be able to see the background 
color `col`.

```{r}
emph <- function(x, col = "bright_yellow", left = "[ ", right = " ]") {
  background_fun <- background(col)
  background_fun(paste0(left, x, right))
}

# Test it out
emph("Hello", "bright_green", "<", ">")
```

## Using dplyr::across(), a word of warning

This section documents a confusing error which can arise from using
`highlight_case()` within `dplyr::across()`.

If you're here because you encountered such an error, the short story is
that the `across()` formula syntax `.fns = ~ f(.x)` is incompatible with
`highlight_case()` and friends.

Consider the case where we want to color all `NA` values blue in the `starwars` 
tibble. The example below shows the what works and what will cause an error. 
The rest of this section goes into detail about what is causing the problem.

```{r, eval = FALSE}
starwars %>%
  mutate(
    across(
      # Using `\()` or `function()` will work
      .cols = where(is_highlightable),
      .fns = \(col) highlight_case(col, is.na(.x) ~ color("blue")(.x))
    
      # Using `~` will raise an error
      # .fns = ~ highlight_case(.x, is.na(.x) ~ color("blue")(.x))
    )
  )
```

### Incorrect output length

vlightr follows the lead of packages across the tidyverse (most 
notably purrr), which use the symbol `.x` to represent an argument to an 
anonymous (AKA lambda) function, e.g. `~ mean(.x, na.rm = TRUE)` in place of 
`\(x) mean(x, na.rm = TRUE)`. While this can make vlightr lambdas more familiar 
to some R users, the symbol `.x` being such a hot commodity does introduce a 
conflict between `dplyr::across()` and `highlight_case()` (as well as vlightr's 
other `*_case()` functions).

Consider the earlier example, where we highlighted the row of any Star
Wars character which is a Droid. With a small change to the `across()`
call, we are faced with a lengthy chained error message.

```{r}
try(
  starwars %>%
  mutate(
    across(
      where(is_highlightable),
      # Before: \(col) templight_case(col, species == "Droid" ~ emph(.x))
      ~templight_case(.x, species == "Droid" ~ emph(.x))
    )
  )
)
```

Reading the end of the message, we see that first formatter 
`attr(,"formatters")[[1]]`, defined by `emph(.x)`, returned a length `r n_rows` 
result (the number of rows in `starwars`) when it should have returned a length
`r n_droids` result (the number of Droids listed in `starwars`), causing
an error.

This happens because, when a formula lambda (e.g.
`~templight_case(...)`) is used within `across()`, the placeholder
symbol `.x` is replaced (usually) with the name of the current column.

For instance, when `name`, the first column in `starwars`, is mutated by
the call to `across()`, the expression `species == "Droid" ~ emph(.x)` 
is substituted with the expression `species == "Droid" ~ emph(name)`.

When `templight_case()` converts the right-hand-side of this formula
into a formatter function, it will look like `function(.x) emph(name)`. This 
formatter will always be in terms of the column `name`, rather than in terms 
of the argument `.x`.

Consequently, when the column `name` is conditionally formatted, the
formatter function receives the names of `r n_droids` Droids to
highlight yellow, but outputs `emph(name)` instead, a length `r n_rows` vector 
of every character name in the `starwars` dataset. This is what causes the 
length error.

We can remedy this error in a several ways:

```{r}
starwars %>%
  mutate(
    across(
      where(is_highlightable),
      # This works:
      \(col) templight_case(col, species == "Droid" ~ emph(.x))
      
      # These also work:
      # ~ templight_case(.x, species == "Droid" ~ \(x) emph(x))
      # ~ templight(.x, species == "Droid", ~ emph(.x))
    )
  ) %>%
  select(1:4) %>%
  head()
```

### Obsolete data mask

A second, even more confusing error, occurs when the `formatters` supplied to 
`templight_case()` produces a result of the correct length. Instead of 
highlighting only droids, let's see what happens when we attempt to highlight 
every column and row.

```{r}
try(
  starwars %>%
    mutate(
      across(
        where(is_highlightable),
        ~templight_case(.x, TRUE ~ emph(.x))
      )
    ) %>%
    format()
)
```

Just like in the previous example, `templight_case()` has created the formatter
function `function(.x) emph(name)` in terms of the column `name`, instead of in
terms of `.x`. Unlike in the previous example, because we're highlighting every
row in `starwars`, this formatter will return a vector of the correct length
(i.e. equal to `nrow(starwars)`) when the highlighted vector is created within 
`across()`.

When the highlighted `starwars` tibble is formatted or printed, however, the 
column `name` no longer exists in the environment of the formatter function 
`function(.x) emph(name)` and thus can't be evaluated, causing the obsolete 
data mask error.

As before, we just need to replace the `~` with `\()` to resolve the issue.

```{r}
starwars %>%
  mutate(
    across(
      where(is_highlightable),
      \(col) templight_case(col, TRUE ~ emph(.x))
    )
  ) %>%
  select(1:4) %>%
  head()
```
