---
title: "Highlighting and dplyr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{highlighting-and-dplyr}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

This vignette describes some functions which are designed for working
with highlighted vectors as columns in a `data.frame` or `tibble` and
provides some examples of manipulating highlights using `dplyr`
functions.

```{r load}
library(vlightr)
library(dplyr, warn.conflicts = FALSE)
```

```{r constants, include = FALSE}
n_droids <- sum(starwars$species == "Droid", na.rm = TRUE)
n_rows <- nrow(starwars)
```

## Using dplyr::across(), a word of warning

This section documents a confusing error which can arise from using
`highlight_case()` within `dplyr::across()` under a specific
circumstance.

If you're here because you encountered such an error, the short story is
that the `across()` formula syntax `.fns = ~ f(.x)` is incompatible with
`highlight_case()` and friends.

Consider the case where we want to colour all `NA` orange in the printed
`starwars` tibble. The example below shows the what works and what will
cause an error. The rest of this section goes into detail about what is
causing the problem.

```{r, eval = FALSE}
starwars %>%
  mutate(
    across(
      # Using `\(col)` or `function(col)` will work
      .cols = everthing(),
      .fns = \(col) highlight_case(col, is.na(.x) ~ color("orange")(.x)))
    
      # Using `~` will raise an error
      # .fns = ~ highlight_case(.x, is.na(.x) ~ color("orange")(.x)))
    )
  )
```

Packages across the tidyverse (most notably purrr) use the symbol `.x`
to represent an argument to an anonymous (AKA lambda) function (e.g.
`~ is.na(.x)`). Because of this, lambdas in vlightr use the same
convention. Unfortunately, the symbol `.x` being such a hot commodity
does introduce one conflict between `dplyr::across()` and
`highlight_case()` (as well as vlightr's other `*_case()` functions).

Consider the earlier example, where we highlighted the row of any Star
Wars character which is a Droid. With a small change to the `across()`
call, we'll introduce an error.

```{r}
try(
  starwars %>%
  mutate(
    across(
      where(is_highlightable),
      # Before: \(col) templight_case(col, species == "Droid" ~ bg("yellow")(.x))
      ~templight_case(.x, species == "Droid" ~ cli::bg_br_yellow(.x))
    )
  )
)
```

The first formatter `attr(,"formatters")[[1]]`, defined by
`cli::bg_br_yellow(.x)`, returned a length `r n_rows` result (the number
of rows in `starwars`) when it should have returned a length
`r n_droids` result (the number of Droids listed in `starwars`), causing
an error.

This happens because, when a formula lambda (e.g.
`~templight_case(...)`) is used within `across()`, the placeholder
symbol `.x` is replaced (usually) with the name of the current column.

For instance, when `name`, the first column in `starwars`, is mutated by
the call to `across()`, the expression
`species == "Droid" ~ cli::bg_br_yellow(.x)` is substituted with the
expression `species == "Droid" ~ cli::bg_br_yellow(name)`.

When `templight_case()` converts the right-hand-side of this formula
into a formatter function, it will look like
`function(.x) cli::bg_br_yellow(name)`. This formatter will always be in
terms of the column `name`, rather than in terms of the argument `.x`.

Consequently, when the column `name` is conditionally formatted, the
formatter function receives the names of `r n_droids` Droids to
highlight yellow, but outputs `cli::bg_br_yellow(name)` instead, a
length `r n_rows` vector of every character name in the `starwars`
dataset. This is what causes the length error.

We can remedy this in a several ways:

```{r}
starwars %>%
  mutate(
    across(
      where(is_highlightable),
      # This works:
      \(x) templight_case(x, species == "Droid" ~ cli::bg_br_yellow(.x))
      
      # These also work:
      # ~ templight_case(x, species == "Droid" ~ \(x) cli::bg_br_yellow(x))
      # ~ templight(x, species == "Droid", ~ cli::bg_br_yellow(.x))
    )
  ) %>%
  select(1:4)
```
