---
title: Conditionally format arbitrary vectors
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{vlightr}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
# TODO: Delete this if error is resolved
#
# Setting `VignetteEngine{rmarkdown::render}` instead of `VignetteEngine{knitr::rmarkdown}`
# because `devtools::build_vignettes()` started failing after I updated `knitr`.
# See: https://forum.posit.co/t/devtools-check-error-with-vignette-but-vignette-can-be-knitted-without-problem/55962/2

knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r}
library(vlightr)
library(lubridate, warn.conflicts = FALSE)
library(ivs)
```

vlightr makes it easy to identify the elements of a vector that you care
about.

```{r}
# Which numbers are above the mean?
above_mean <- highlight(1:10, ~ .x > mean(.x), ~ paste0("[", .x, "]"))
print(above_mean)
```

Including observations in a tibble.

```{r, collapse=TRUE}
tibble::tibble(x = above_mean[1:5])
```

Highlights are generic, meaning that many vectors you'll encounter in
the wild can be highlighted. For example, here's a lubridate interval
vector with a festive format method.

```{r}
ints <- lubridate::interval(
  start = ymd("2023-12-01") + ddays(c(0, 6, 14)),
  end = ymd("2023-12-01") + ddays(c(2, 8, 26))
)

format_festive <- function(x) {
  out <- strsplit(x, "")[[1]]
  evens <- seq_along(out) %% 2 == 0
  out[evens] <- cli::col_red(out[evens])
  paste(cli::col_green(out), collapse = "")
}

highlight(ints, ~ ymd("2023-12-25") %within% .x, ~ sapply(.x, format_festive))
```

### Conditional Formatting

For it's basic functionality, the `highlight()` function needs 3
arguments: `x`, `conditions`, and `formatters`.

```{r, eval = FALSE}
above_mean <- highlight(
  # A vector to conditionally format
  x = 1:10,
  # Condition(s), determine which elements of `x` are formatted
  conditions = ~ .x > mean(.x),
  # Formatter(s), specify how chosen elements of `x` are formatted
  formatters = ~ paste0("[", .x, "]")
)
```

Multiple `conditions` and `formatters` can be specified by supplying
lists of functions or purrr-style lambdas. Elements of `x` for which
`conditions[[i]](x)` returns `TRUE` are formatted using
`formatters[[i]]`.

```{r}
even_or_negative <- highlight(
  x = -4:3,
  conditions = list(
    \(x) x %% 2 == 0,        # test if `x` is even
    \(x) x < 0               # test if `x` is negative
  ),
  formatters = list(
    ~ paste0("[", .x, "]"),  # format if `x` is even
    ~ paste0("(", .x, ")")   # format if `x` is negative
  )
)
print(even_or_negative)
```

By default, `formatters` are applied in the order that they were
supplied. You can change the order of application by supplying a
`precedence`.

```{r}
# Change even and negative format from "([])" to "[()]"
highlight(
  x = even_or_negative,
  conditions = list(\(x) x %% 2 == 0, \(x) x < 0),
  formatters = list(~ paste0("[", .x, "]"), ~ paste0("(", .x, ")")),
  precedence = c(2, 1)
)
```

A highlighted vector's conditional formatting can also be updated using
`update_highlight()`, which takes the same arguments as `highlight()`.
The following is equivalent to the update made above.

```{r}
even_or_negative <- update_highlight(even_or_negative, precedence = c(2, 1))
print(even_or_negative)
```

Elements of `even_or_negative` which meet multiple conditions (i.e. are
even and negative) are formatted multiple times (e.g. `"[(-4)]"`,
`"[(-2)]"`). To format elements according to the first condition met,
set `format_once = TRUE`.

```{r}
update_highlight(even_or_negative, format_once = TRUE)
```

If it's unclear how a highlighted vector is being formatted, use `describe_highlight()`
to get a brief description of a highlight's conditional formatting.

```{r}
describe_highlight(even_or_negative)
```

The `describe_highlight()` function uses the `description` attribute of
a highlighted vector to provide additional context.

```{r}
even_or_negative <- update_highlight(
  even_or_negative, 
  description = c("Even", "Negative")
)
describe_highlight(even_or_negative)
```

### Manipulating Highlighted Data

Highlighted vectors (class `vlightr_highlight`) are designed to
introduce as little friction as possible, but they do require some
legwork to use. Attempting to coerce or convert a `vlightr_highlight` to
another vector type will result in an error.

```{r}
mean_or_above <- highlight(1:6, ~ .x >= mean(.x), ~ paste0("[", .x, "]"))
try(mean_or_above[[5]] <- 100L)
```

Highlights are, however, compatible with other highlights. Where
possible, highlighted vectors have the same casting and coercion rules
as their underlying vector type. `hl()` is a synonym for `highlight()`,
which is useful for quick inline highlighting.

```{r}
# <highlight<integer>> and <highlight<integer>> are compatible
mean_or_above[[5]] <- hl(100L)

# <highlight<integer>> and <highlight<character>> are not
try(mean_or_above[[5]] <- hl("100!"))
```

Arithmetic, comparison, and the math functions supported by
`vctrs::vec_math()` work out of the box.

```{r}
# Arithmetic
mean_or_above - hl(2.0)
# Comparison
pmin(mean_or_above, hl(4))
# Math
sum(mean_or_above)
```

Notice that the conditional formatting is persistent, so changing the
underlying data of a highlighted vector will also change how each
element is formatted.

### Highlight, Un-Highlight, Re-Highlight

By default, many functions won't know what to do with a highlighted
vector, resulting in an error.

```{r}
try(as.logical(mean_or_above))
try(which.min(mean_or_above))
```

A highlight can be un-highlighted (converted back to it's original
class) using the `un_highlight()` function. This allows the highlighted
vector to be used as expected, at the expense of it's conditional
formatting.

```{r}
as.logical(un_highlight(mean_or_above))
which.min(un_highlight(mean_or_above))
```

The `re_highlight()` function can be used to re-apply the conditional
formatting removed by `un_highlight()`.

```{r}
re_highlight(as.logical(un_highlight(mean_or_above)), mean_or_above)
re_highlight(which.min(un_highlight(mean_or_above)), mean_or_above)
```

Together, `highlight()`, `un_highlight()`, and `re_highlight()` provide
the core functionality of `vlightr`.

1.  `highlight()` (synonym `hl()`) highlights a vector.
2.  `un_highlight()` (synonym `ul()`) removes a vector's highlighting to expose it's underlying data.
3.  `re_highlight()` (synonym `rl()`) applies the format of an existing
    highlighted vector to another vector.

These functions and their shorthand synonyms allow highlights to be used
almost as if they were normal vectors.

```{r}
# Emphasize `NA` values
indicators <- highlight(c(1L, 0L, NA, 1L), is.na, ~ paste("[", .x, "]"))
indicators
# Convert to a logical vector
indicators |>
  un_highlight() |>
  as.logical() |>
  re_highlight(indicators)
```

The "highlight-pipe" `%hl>%` is a wrapper around the
[magrittr](https://magrittr.tidyverse.org/reference/pipe.html) `%>%`
pipe which un-highlights the left-hand-side input and re-highlights the
right-hand-side output.

```{r}
indicators %hl>% as.logical()
```

`re_highlight(x, ...)` accepts any number of highlighted vectors `...`
which are used to re-highlight the vector `x`. This is useful for
restoring the conditional formatting of multiple highlights used within
a single function.

```{r}
# Highlight phrases containing an exclamation
exclamations <- "(wow|neat|cool)"
exclaim <- highlight(
  c("wow that's great", "okay", "what a cool dude", "no"),
  ~ grepl(exclamations, .x),
  ~ paste0(.x, "!")
)
exclaim
# Highlight phrases containing a question
questions <- "(what|where|when)"
question <- highlight(
  c("news", "neat, when is that", "where is he", "thanks"),
  ~ grepl(questions, .x),
  ~ paste0(.x, "?")
)
question
# Un-Highlight, Do Something, Re-Highlight
paste(ul(exclaim), ul(question)) |> rl(exclaim, question)
```

If the `x` argument to `re_highlight()` is a highlighted vector, then
it's conditional formatting and that of the highlights supplied to `...`
will be combined. Only the data from `x` is used in the output vector.

```{r}
re_highlight(x = exclaim, question)
```

### Applying Multiple Formats

Multiple conditions and formatters can be specified by supplying lists
of functions or purrr-style lambdas.

```{r}
min_max <- highlight(
  x = 1:5,
  conditions = list(
    ~ .x == min(.x), 
    ~ .x == max(.x)
  ),
  formatters = list(
    ~ paste0("(Min = ", .x, ")"), 
    ~ paste0("(Max = ", .x, ")")
  )
)
print(min_max)
```

When two compatible highlights are combined, so are their `conditions`
and `formatters`.

```{r}
c(min_max[1:3], mean_or_above[2:4])
```

An error will occur if the combination of multiple highlights causes any
of `conditions` or `formatters` to fail.

```{r}
# A highlighted vector with a hidden error
int_hl <- highlight(
  c(1L, 2L, 3L), 
  conditions = ~ !is.integer(.x),
  formatters = \(x) stop("Integers only!")
)
```

```{r}
# This runs fine
int_hl
# This does not
try(c(int_hl, hl(0.5)))
```

The `precedence` of highlighted vectors is maintained when they are
combined. For example, you can set a highlighted vector's `precedence`
to prioritize it's format.

```{r}
even_is_red <- hl(1:3, ~ .x %% 2 == 0, cli::col_red, precedence = 2)
even_is_green <- hl(4:6, ~ .x %% 2 == 0, cli::col_green, precedence = 1)
c(even_is_red, even_is_green)
```

When defining more complex conditional formatting, `highlight_case()`
(inspired by `dplyr::case_when()`) can make the relationship between a
condition and it's corresponding format more clear.

```{r}
summary_hl <- highlight_case(
  x = 1:5,
  .x == min(.x) ~ paste0("(Min = ", .x, ")"),
  .x == quantile(.x, 0.5) ~ paste0("(Mid = ", .x, ")"),
  .x == max(.x) ~ paste0("(Max = ", .x, ")"),
  .x > quantile(.x, 0.5) ~ paste("[ High", .x, "]"),
  .x < quantile(.x, 0.5) ~ paste("[ Low", .x, "]")
)
summary_hl
```

Unlike in `highlight()`, the `.format_once` argument of
`highlight_case()` is `TRUE` by default. As in `dplyr::case_when()`,
this means that each element of `x` is allowed to match exactly one
condition (the first condition met).

### Highlighters

`as_highlighter()` converts a conditionally formatted vector to a
function which applies the same conditional formatting to other vectors.

```{r}
summary_highlighter <- as_highlighter(summary_hl)
summary_highlighter(-2:2)
```

Highlighter functions accept the same arguments as `highlight()`, so
additional highlights can be applied on top of existing ones.

```{r}
summary_highlighter(
  x = -2:2, 
  conditions = ~ .x > 0, 
  formatters = ~ paste0("+", .x),
  precedence = 0,
  format_once = FALSE
)
```

A highlighter can be defined from scratch using `highlighter()` or
`highlighter_case()`. These functions take the same arguments as
`highlight()` and `highlight_case()` respectively, excluding the first
argument `x`.

```{r}
secret <- highlighter(
  conditions = ~ grepl("^<.*>$", .x),
  formatters = ~ strrep("x", nchar(.x) - 2L)
)
secret(c("This", "<is>", "a", "<secret>", "<.>"))
```

```{r}
dollars <- highlighter_case(
  is.na(.x) ~ "NA",
  .x > 0 ~ paste0("-$", .x),
  TRUE ~ paste0("$", .x),
  .init_formatter = \(x) sprintf("%.2f", abs(x))
)
dollars(c(-10, 0.12, 130.0234, NA))
```

### Styling

Highlights can be formatted using ANSI escape sequences, such as those
used to color text in the cli package. The code below colors the
formatted `NA` values red.

```{r}
highlight(c(1, 0, NA), is.na, cli::col_red)
```

vlightr provides some convenience functions for selecting cli functions
to use as formatters.

-   `color()` and `colour()` return a `cli::col_*` function which colors
    text

-   `bg()` and `background()` return a `cli::bg_*` function which colors
    the text's background

-   `style()` returns a `cli::style_*` function styles text (e.g. underlines,
    italicizes, etc.)

For example, the same `NA` highlight as above can be achieved like so:

```{r}
highlight(c(1, 0, NA), is.na, color("red"))
```

See `cli::style_bold()` and friends for details on ANSI colored and
styled text.

### Highlighting Vectors in the Wild

Informally, a highlight-able vector is some collection of size 1
elements. This means that a list, which can contain elements of varying sizes, 
is non-highlightable (although a list can contain highlighted vectors).

Formally, `is_highlightable()` uses the following criteria to test
whether an object `x` can be highlighted:

-   `x` is not a `data.frame()` (or dataframe subclass)

-   `x` is not a bare `list()`

-   `x` is otherwise a vector, as defined by `vctrs::obj_is_vector()`

The `highlight()` function (and friends `hl()`, `highlight_case()`, `hl_case()`) 
provide some extra arguments which are useful for refining how user-generated 
vectors are highlighted.

Let's highlight a `lubridate::interval` vector named `long_interval` and refine 
it's format method using these extra arguments.

```{r}
long_interval <- interval(
  start = ymd_hms("19910101120000") + dyears(c(5, 10, 15)), 
  end = ymd_hms("20240101120000") + dseconds(c(3, 7, 12)), 
  tzone = "UTC"
)

highlight(long_interval)
```

The `init_formatter` ("initial formatter") is a function which is called at the 
start of a highlighted vector's `format()` method, before any conditional formatting 
takes place. It expects a single input, the highlighted vector's data, and should 
return a formatted vector as it's result.

Suppose we don't want to see the hours-minutes-seconds component of the dates in
our `long_interval`. We can define an initial formatter `format_interval` to do
just that. While we're at it, we can remove the first timezone (`"UTC"`) from each
element of the formatted interval

```{r}
format_interval <- function(x) {
  # Remove the h-m-s component of the interval's start and end
  start <- as.Date(int_start(x))
  end <- as.Date(int_end(x))
  
  # Format the output as "start--end tzone"
  out <- paste0(start, "--", end, " ", attr(x, "tzone"))
  
  # Remember to preserve NA values from `x`
  out[is.na(x)] <- NA_character_
  out
}

highlight(long_interval, init_formatter = format_interval)
```

The `last_formatter` is a function which is called at the end of a highlighted 
vector's `format()` method, after it has been conditionally formatted. This is 
useful for limiting the text width of a formatted highlight.

```{r}
# Truncate elements of a character vector such that `all(nchar(x) <= char_limit)`
format_truncated <- function(x, char_limit = 40) {
  # Check which strings in `x` exceed the `char_limit`
  x_char <- nchar(x)
  above_limit <- x_char > char_limit
  if (!any(above_limit)) {
    return(x)
  }
  
  # Split strings above the `char_limit` into a prefix and suffix, 
  # leaving room to insert a "..." in between.
  substr_size <- floor(char_limit / 2)
  suffix_start <- x_char[above_limit] - substr_size + 2
  x_prefix <- substr(x[above_limit], 1, substr_size - 1)
  x_suffix <- substr(x[above_limit], suffix_start, x_char[above_limit])
  
  # Truncate the too-long strings and insert a "..." in the middle
  x[above_limit] <- paste0(x_prefix, "...", x_suffix)
  x
}

# Test the formatter
format_truncated(c("this", "is", "a", "huuuuuuuuuge", "character"), 7)
```

After packaging our custom `init_formatter` and `last_formatter` together in a 
highlighter, we can reuse them wherever we highlight an interval vector.

```{r}
highlight_interval <- highlighter(
  init_formatter = format_interval,
  last_formatter = format_truncated
)

highlight_interval(
  long_interval, 
  ~ym("200001") %within% .x, 
  ~paste(.x, "[Contains the Millenium!!]")
)
```

### Under the Hood

This section is meant to provide some intuition into how
`vlightr_highlight` vectors are formatted.

Consider the vector `x_hl` defined below. We'll just assume that `x` is
some highlight-able vector and `conditions`, `formatters`, `precedence`,
`init_formatter`, and `last_formatter` are all valid arguments.

```{r, eval = FALSE}
x_hl <- highlight(
 x = x,
 conditions = conditions,
 formatters = formatters,
 precedence = precedence,
 init_formatter = init_formatter,
 last_formatter = last_formatter
)
```

When `format(x_hl)` is called (e.g. within `print(x_hl)`) the following
pseudo-code is executed.

```{r, eval = FALSE}
# Apply an initial format to `x`, converting `x` into a character vector.
formatted <- init_formatter(x)

# Apply each format in `formatters` to the elements of `x` which meet a 
# condition. Do so in the order specified by `precedence`.
format_order <- order(precedence)
for (i in format_order) {
  condition <- conditions[[i]]
  formatter <- formatters[[i]]
  condition_is_met <- condition(x)
  formatted[condition_is_met] <- formatter(formatted[condition_is_met])
}

# Apply the last format before returning
last_formatter(formatted)
```

The `init_formatter` is the only formatter function which has access to
the underlying vector data in `x`. As such, all conditional formatting
performed by the `conditions` or `formatters` could be performed by the
`init_formatter` alone.
